---
title: "Android - Button Click"
author: Laxmi K Soni 
description: "Handling Button click in Android"
slug: Handling Button click in Android
date: 2018-01-05
lastmod: 2018-01-05
categories: ["Mobile Application Development"]
tags: ["Mobile Application Development"]
Summary: Handling Button click in Android
subtitle: Handling Button click in Android
featured: "img/main/2018-01-05-Android-Button-Click-1.jpg"
output:
  blogdown::html_page:
    toc: false
  html_document:
    highlight: tango
    theme: flatly
    toc: false
---


```{r setup, include=FALSE}
library(tidyverse)
library(magick)
library(reticulate)
library(Rcpp)


conda_list()[[1]][1] %>% 
  use_condaenv(required = TRUE)
use_python("C://Users//slaxm//CONDA~1//envs//myenv1//python.exe")


```

#### Using anonymous Inner class


```{r,echo=TRUE , eval=FALSE}
((Button)findViewById(R.id.random_numbers_btn)).setOnClickListener(new View.OnClickListener() {
    @Override
    public void onClick(View v) {
        Random r = new Random();
        int i = r.nextInt(100 - 1) + 1;
        ((TextView)findViewById(R.id.random_numbers_tv)).setText(String.valueOf(i));
    }
});
```
 
#### Using an instance of the Button.onClickListener

```{r,echo=TRUE , eval=FALSE}
Button.OnClickListener but_listener = new Button.OnClickListener() {
      @Override
      public void onClick(View v) {
          Random r = new Random();
          int i = r.nextInt(100 - 1) + 1;
          ((TextView)findViewById(R.id.random_numbers_tv)).setText(String.valueOf(i));
      }
  };
  
((Button)findViewById(R.id.random_numbers_btn)).setOnClickListener(but_listener);
```


#### Using onClick attribute of the button in the layout xml file

```{r,echo=TRUE , eval=FALSE}
    <Button
        android:id="@+id/random_numbers_btn"
        android:layout_width="wrap_content"
        android:layout_height="wrap_content"
        android:text="Generate Random Numbers"
        app:layout_constraintEnd_toEndOf="parent"
        app:layout_constraintStart_toStartOf="parent"
        tools:layout_editor_absoluteY="185dp"
        tools:ignore="MissingConstraints"
        android:onClick="generateRandomNumbers"
        />

   public void generateRandomNumbers(View view) {
        Random r = new Random();
        int i = r.nextInt(100 - 1) + 1;
        ((TextView)findViewById(R.id.random_numbers_tv)).setText(String.valueOf(i));
    }
        
```


#### Using class implementing the listner

```{r,echo=TRUE , eval=FALSE}

public class MainActivity extends AppCompatActivity implements View.OnClickListener {

    @Override
    protected void onCreate(Bundle savedInstanceState) {
        super.onCreate(savedInstanceState);
        setContentView(R.layout.activity_main);
        ((Button)findViewById(R.id.random_numbers_btn)).setOnClickListener(this);

    }

    public void generateRandomNumbers(View view) {
        Random r = new Random();
        int i = r.nextInt(100 - 1) + 1;
        ((TextView)findViewById(R.id.random_numbers_tv)).setText(String.valueOf(i));
    }

    @Override
    public void onClick(View v) {
        generateRandomNumbers(v);
    }
}
```

#### Using annotations and ButterKinfe

- replace findViewById calls by using @BindView on fields.

- Group many views in a list or array. Operate on all of them at once with actions, setters, or properties.

- replace anonymous inner-classes for listeners by annotating methods with @OnClick and others.

- replace resource lookups by using resource annotations on fields.



```{r,echo=TRUE , eval=FALSE}

android {
  ...
  // Butterknife requires Java 8.
  compileOptions {
    sourceCompatibility JavaVersion.VERSION_1_8
    targetCompatibility JavaVersion.VERSION_1_8
  }
}

dependencies {
  implementation 'com.jakewharton:butterknife:10.2.3'
  annotationProcessor 'com.jakewharton:butterknife-compiler:10.2.3'
}
```

```{r,echo=TRUE , eval=FALSE}

buildscript {
    repositories {
        google()
        jcenter()
    }
    dependencies {
        classpath "com.android.tools.build:gradle:4.1.2"
        classpath 'com.jakewharton:butterknife-gradle-plugin:10.2.3'
        // NOTE: Do not place your application dependencies here; they belong
        // in the individual module build.gradle files
    }
}

allprojects {
    repositories {
        google()
        jcenter()
    }
}

task clean(type: Delete) {
    delete rootProject.buildDir
}

```


```{r,echo=TRUE , eval=FALSE}

public class Main_Activity extends Activity {

    protected void onCreate(Bundle savedInstanceState) {
        super.onCreate(savedInstanceState);
        setContentView(R.layout.activity_main);
        ButterKinfe.inject(this); 
    }

    @OnClick(R.id.button_id)
    void onButtonClicked() {
    }
    

    @OnClick(R.id.other_button_id)
    void onOtherButtonClicked(View otherButton) {
    }

}
```


#### Summary:

In android, Button is a user interface control that is used to perform an action whenever the user clicks on it.
Generally, Buttons in android will contain a text or an icon or both and perform an action when the user touches it.